---
title: "Just in Case Copy"
author: "Hailey Knowles"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

---
title: "Final Project"
author: "Hailey Knowles"
date: "`r Sys.Date()`"
output:
  github_document:
    toc: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# ABSTRACT

This research project investigated the effect of location (Northern vs. Southern counties in Utah) on the abundance of fireflies observed. The data was collected from various locations, and the counts of fireflies were compared by region using visualization analysis (box plot and violin plot) and also using a quasipoisson regression, which is used when the data collected is over-dispersed. The analysis indicated a significant difference in firefly abundance across regions. However, the region with the higher abundance was the opposite of the original hypothesis. The Southern counties in Utah showed a higher abundance of fireflies compared to the Northern counties. After examining these findings, it is important to understand the importance of choosing appropriate statistical models and tests for ecological data, which suggests that the region in Utah may affect firefly abundance.

# BACKGROUND

Fireflies are a fascinating species that many researchers are interested in due to their glowing abilities. They are classified in the Lampyridae family of insects. According to Firefly Conservation and Research, there are over 2400 firefly species in 144 genera (Firefly Conservation and Research, 2025). According to a study conducted in 2023, it was determined that water is a vital source for fireflies and is required for them to thrive in their various environments. Fireflies have a preference for areas with high humidity and constant moisture. They also prefer areas with abundant trees and various types of grasses (Qing & Lin, 2023). Fireflies need to lay their eggs in moist soil or even in damp tree trunks. Many fireflies are found in tropical regions, but there are a few fireflies that can be found in dry forests or arid environments, but only during the rainy seasons. These fireflies can be found in every part of the world except for Antarctica. These kinds of conditions are more common in the Northern areas of Utah compared to the Southern areas of Utah (Grimaldi, 2025).

There is a wide variety of climates within the state of Utah, and some of these variations are based on the Northern and Southern areas. The further South you move, the less rainfall you see and the smaller amount of green vegetation. Utah has an arid climate with little annual rainfall, receiving less than 15 inches of rain per year. It was found that the Northern areas of Utah tend to be cooler on average than the Southern Utah regions. During the whole year, there is a tendency for the Northern regions of Utah to also receive more rain than the Southern regions (GottAgoUtah_admin, 2024).

# STUDY QUESTION and HYPOTHESIS

## Question

Does the location (Northern or Southern counties) of the fireflies affect the level of abundance of fireflies in Utah?

## Hypothesis

We hypothesize that Northern counties in Utah will have a higher overall abundance compared to Southern counties due to a lower overall temperature in the Northern counties and a higher abundance of green vegetation and rainfall when compared to the Southern counties.

## Prediction

We predict that Northern counties in Utah will have a higher overall abundance compared to Southern counties. A possible test statistic is a quasipoisson regression model. This would allow us to compare the two variables of north and south abundance values to determine which one has a higher abundance and if the difference is significant.

# METHODS

The data that was used for this research project was collected from a firefly observation project conducted in various locations around Utah. The exact locations of the observations were hidden due to the protection of privacy. The data was then organized by county, abundance count, date, and had individualized notes about where the fireflies were seen and the type of environment.

The data sheet that was filled out and organized by our research team allowed the running of various tests and the creation of plots. Each county in Utah within the data set was assigned to be either Northern or Southern based on distinctions made by our research team on what constituted North or South. The distinction made was that any county that was in line with Carbon County or lower would be considered a Southern county, and anything above would be considered a Northern county.

The plots created based on the data were a box plot and a violin plot. The box plot allows us to visualize the data in a way to compare the mean values and distribution for both North and South. The other plot is a violin plot, which allows the visualization of the distribution of the abundance of fireflies based on their location in the North or South counties. The final statistical analysis that was run was a quasipoisson regression, which allowed the test to assume that the data distribution was not normal and had a skewed tail. It was determined that the data was over dispersed and therefore allowed the use of the quasipoisson regression because the variance was greater than the mean. It compares all the values in the data set and determines if the difference is significant or not significant.

## 1st Analysis - Box Plot

This is a box plot that allows the visualization of the various aspects of the data set, comparing the abundance of fireflies in Northern and Southern counties in Utah. It allows the visual comparison of the mean, max, min, and outliers.

```{r}
# Firefly Boxplot (Log-Transformed, No Blanks)

library(ggplot2)

# Read in the data
fireflies <- read.csv("Copy of firefliesUtah - Usable Data.csv", stringsAsFactors = FALSE)
colnames(fireflies) <- c("firefly_count", "region")

# Remove blank or missing region values
fireflies <- subset(fireflies, region != "" & !is.na(region))

# Box plot with log10 transformation (+1 to avoid log(0))
ggplot(fireflies, aes(x = region, y = log10(firefly_count + 1), fill = region)) +
geom_boxplot(width = 0.6, color = "black", alpha = 0.7) +
labs(
title = "Firefly Abundance by Region (Log Scale)",
x = "Region",
y = "Log-Transformed Firefly Count"
) +
scale_fill_manual(values = c("north" = "#00A9FF", "south" = "orange")) +
theme_minimal(base_size = 13) +
theme(
legend.position = "none",
plot.title = element_text(size = 23, face = "bold", hjust = 0.5),
axis.text = element_text(size = 20, face = "bold"),
axis.title = element_text(size = 20, face = "bold"),
panel.grid.major.x = element_blank(),
panel.grid.minor = element_blank()
)
#ChatGPT. OpenAI, version GPT-5.1. Used as a first pass, but edited carefully to make the figure look correct. Accessed [November 2025].
```

## 2nd Analysis - Violin Plot

This is a violin plot that expresses the data in a visual way to compare the distributions of the firefly abundance in Northern and Southern counties. The distribution is important to understand and look at side-by-side because then you can see what is the most common number of abundance observations.

```{r}
# Split Violin Plot (y-axis limited to 50) and downloaded the necessary library packages

library(ggplot2)
library(gghalves)
library(stringi)

# Read and cleaned the data
fireflies <- read.csv("Copy of firefliesUtah - Usable Data.csv", stringsAsFactors = FALSE)
colnames(fireflies) <- c("firefly_count", "region")

fireflies$region[fireflies$region == ""] <- NA
fireflies$region <- stri_trans_general(fireflies$region, "NFKC")
fireflies$region <- stri_replace_all_regex(fireflies$region, "\\p{C}", "")
fireflies$region <- gsub("\u00A0", " ", fireflies$region)
fireflies$region <- trimws(tolower(fireflies$region))
fireflies$region[fireflies$region %in% c("n", "nrth", "noth")] <- "north"
fireflies$region[fireflies$region %in% c("s", "sth", "soth")] <- "south"
fireflies$region <- factor(fireflies$region, levels = c("north", "south"))
fireflies_clean <- droplevels(subset(fireflies, !is.na(region)))

# Split the violin plot
ggplot() +
geom_half_violin(
data = subset(fireflies_clean, region == "north"),
aes(x = factor(1), y = firefly_count, fill = region),
side = "l", trim = TRUE, color = "black", alpha = 0.7
) +
geom_half_violin(
data = subset(fireflies_clean, region == "south"),
aes(x = factor(1), y = firefly_count, fill = region),
side = "r", trim = TRUE, color = "black", alpha = 0.7
) +
scale_fill_manual(values = c("north" = "#00A9FF", "south" = "orange")) +
coord_cartesian(ylim = c(0, 50)) + # y-axis capped at 50
labs(
title = "Firefly Abundance: North vs South",
x = NULL,
y = "Firefly Count"
) +
theme_minimal(base_size = 13) +
theme(
legend.position = "bottom",
plot.title = element_text(size = 24, face = "bold", hjust = 0.5),
axis.text.x = element_blank(),
axis.ticks.x = element_blank(),
axis.title.y = element_text(size = 20, face = "bold"),
axis.title.x = element_text(size = 20, face = "bold")
)
#ChatGPT. OpenAI, version GPT-5.1. Used as a first pass, but edited carefully to make the figure look correct. Accessed [November 2025].
```

## 3rd Analysis - Wilcoxon Test

This is a quasipoisson regression analysis that compares the values of firefly abundance in the two groups (Northern and Southern counties in Utah) to see if they are statistically different from each other. In this case, with a p-value of 0.00344, we can understand that there is a statistically significant difference between these two groups, with the Southern counties being of higher value.

```{r}
# Shapiro–Wilk tests for normality

shapiro.test(
  fireflies$firefly_count[fireflies$region == "south"]
)
## 
##  Shapiro-Wilk normality test
## 
## data:  fireflies$firefly_count[fireflies$region == "south"]
## W = 0.19196, p-value < 2.2e-16
shapiro.test(
  fireflies$firefly_count[fireflies$region == "north"]
)
```



```{r}
wilcox.test(
  firefly_count ~ region,
  data = fireflies
)
```

Wilcoxon Test Accounting for greater only
```{r}
wilcox.test(firefly_count ~ region, data = fireflies,
            alternative = "greater")
```

```{r}
tapply(fireflies$firefly_count, fireflies$region, median)
```



# DISCUSSION

Given the results from the box plot, violin plot, and the quasipoisson regression, we can conclude that there is a statistically significant difference between the abundance of fireflies observed in the Northern and Southern counties in Utah.

## Interpretation of 1st Analysis (Box Plot)

The box plot shows information comparing the means, minimums, maximums, and the first and third quartiles. While this information alone cannot fully support our hypothesis, it indicates that Northern counties in Utah do not have a higher abundance of fireflies than Southern counties. The visual comparison shows that the two groups' means are similar, but the Southern counties have slightly higher median values.

## Interpretation of 2nd Analysis (Violin Plot)

The violin plot provides another visual representation of the data. It allows the observation counts to be compared side-by-side for Northern and Southern counties. In the plot, there is a cluster of higher abundance around four fireflies per observation in the Southern counties. The Northern counties show high spikes in abundance, making the counts less consistent, while the Southern counties show a more gradual, consistent increase. This supports the idea that Southern counties may have more stable firefly populations compared to the Northern region.

## Interpretation of 3rd Analysis (Quasipoisson Regression)

The quasipoisson regression confirms a significant difference between the two groups. The p-value in the analysis was 0.00344, indicating a statistically significant difference in abundance between Northern and Southern counties. Specifically, the Southern counties had higher abundance counts than the Northern counties.

The quasipoisson model was appropriate because the data were not normally distributed, and both the Northern and Southern regions showed long-tailed distributions. These high-count observations lengthened the tail and influenced the distribution even when the averages mainly remained moderate. The Northern region contains some abnormally high-count observations, while the Southern region shows more consistent patterns of abundance.

# CONCLUSION

Based on the evidence collected, there is no support for the original hypothesis that predicted a higher overall observation abundance in the Northern counties in Utah compared to the Southern counties. Each analysis conducted showed that there is a significant difference between the mean abundance measurements and that a significant conclusion can be drawn, but in the opposite direction from the initial hypothesis. The Southern counties were found to have a statistically significantly higher number of observations compared to the Northern counties in Utah. 

There were many limitations that occurred during this experiment. This data was collected from various locations in Utah, but the exact location of the sightings was undisclosed and may not be exact. Another uncertainty involves the way the data was collected. Observations were collected by multiple observers and counted manually, which may not be exactly accurate since many of the observation numbers should be considered approximate due to the difficulty of actually counting flying fireflies. The counts in Northern and Southern counties might be influenced by the Northern counties having many more observation counts compared to the Southern counties' observations. The data used in this experiment fail to support the original hypothesis, and future studies should aim to determine the differences between climates and their impact on firefly observation counts.

# REFERENCES

1.  ChatGPT. OpenAI, version Jan 2025. Used as a reference for functions such as plot() and to correct syntax errors. Accessed `r Sys.Date()`.

2.  Firefly Life Cycle & Habitat: Lightning bug facts. Orkin. (n.d.). <https://www.orkin.com/pests/be>

3.  GottAgoUtah_admin. (2024, December 9). Understanding Utah’s year-round weather conditions. Gotta Go Utah ! <https://www.gottagoutah.org/understanding-utahs-year-round-weather-conditions/>

4.  Shen, M., Qing, Z., & Lin, S. (2023). The Impact of Environment Situation on Fireflies and the Contribution of Fireflies on Environment Situation.

5.  Types of fireflies. Firefly Conservation & Research. (2023, March 27). <https://www.firefly.org/types-of-fireflies.html>

6.  West, F. L. R., & Edlefsen, N. E. (1919). The climate of Utah (No. 166). Utah Agricultural College Experiment Station.
